Informing IPCC accounting of forest carbon using the global forest carbon database (ForC v4.0)

### 2022-04-11 doc/manuscript/ESSDmanuscript.Rmd
work on intro

bibliography: references.bib
running:
  title: IPCC forest C accounting with ForC
  author: Williams et al.
competinginterests: |
  The authors declare no competing interests.
# OPTIONAL:
algorithms: true
# See https://publications.copernicus.org/for_authors/licence_and_copyright.html, normally used for transferring the copyright, if needed. 
# Note: additional copyright statements for affiliated software or data need to be placed in the data availability section. 
copyrightstatement: |
  The author's copyright for this publication is transferred to institution/company. 
### Note: unless stated otherwise, software and data affiliated with the manuscript are assumed to be published under the same licence as the article (currently Creative Commons 4.0)
availability:
  #  use this to add a statement when having only software code available
  #  use this to add a statement when having only data sets available
  #code: |
  #data: |
  codedata: |
    use this to add a statement when having data sets and software code available
authorcontribution: |
  (fill this in)
disclaimer: |
acknowledgements: |
  Thank you to all researchers who collected and published the data contained in ForC, and to all research assistants and collaborators who have helped to build the database.
  Funding for this study was provided by Bezos Earth Fund to The Nature Conservancy, the Institute for Global Environmental Strategies, WLS(?)
appendix: |
   \section{Mapping ForC to EFDB}

@@

Reflecting their strong influence on Earth's climate, forests play a central role in international plans for climate change mitigation under the Paris Agreement [@unfccc_adoption_2015].
Forest conservation, reforestation, and improved sustainable management all have significant -- and relatively cost-effective -- potential as climate change mitigation options, with conservation and reforestation having the fourth and fifth largest net emission reduction potentials or all mitigation options [@ipcc_summary_2022].
As of 2016, forest-based mitigation accounted for 26% of total planned greenhouse gas mitigation within Nationally Determined Contributions under the Paris Agreement [@grassi_key_2017].
Yet, envisioned forest-based climate change mitigation initiatives do not always correspond to actual emission reductions through on-the-ground implementation [@refs].
One critical need for ensuring that forest-based climate change mitigation initiatives are effective is realistic planning, underlain by solid scientific data [@anderson-teixeira_effective_2022].

The International Panel on Climate Change (IPCC) provides guidance for national greenhouse gas inventories for reporting to the United Nations Framework Convention on Climate Change [UNFCCC, @ipcc_2019_2019].
Under this guidance, greenhouse gas inventories include all managed land, including most of the world's forest land [@ogle_delineating_2018].
The IPCC inventory guidelines include specific instructions for accounting for greenhouse gas (mainly CO~2~) exchanges between forest land and the atmosphere [@ipcc_agriculture_2006; @ipcc_2019_2019].
This guidance has improved over the years as more of the relevant underlying data has become available [@requenasuarez_estimating_2019; @rozendaal_aboveground_2022], but there remains room for continuous improvement as the science advances.
For example, the year following the release of the latest IPCC guidelines, @cook-patton_mapping_2020 found that the latest default rates may underestimate rates of C accumulation in regrowth forests by 32% on average and fail to capture eight-fold variation within ecozones. 
In addition, @cuni-sanchez_high_2021 found that aboveground C stocks in mature African tropical montane forests were two-thirds higher than the IPCC default values for these forests.
This rapid evolution of scientific information on the climate mitigation potential of forests is beneficial to climate mitigation efforts, but requires improved mechanisms for communicating the latest information from scientific researchers to the practitioners who need reliable estimates for greenhouse gas mitigation planning. 
Moreover, high variability of forest C cycling within ecozones [e.g., @cook-patton_mapping_2020; @refs] implies that it is useful for those compiling national greenhouse gas inventories to have access to locally-specific information, when available.
To improve the data accessible for C accounting, the IPCC created the Emission Factor Database (EFDB; https://www.ipcc-nggip.iges.or.jp/EFDB/main.php), which is intended as a recognized library of emission factors and other parameters that can be used for estimating greenhouse gas emissions and removals.

**To ensure that planned emissions reductions are realistic, high-quality estimates of forest C stocks and fluxes must be publicly accessible.**
The EFDB is intended as a recognized library, and can be used both for efforts to tally a nation's intended or accomplished greenhouse gas reductions, or as a basis of comparison for external parties to evaluate these inventories. 
<!-- check Washington Post article on Malaysia NDC, @deng_comparing_2021 -->

>>>>

bibliography: references.bib
running:
  title: IPCC forest C accounting with ForC
  author: Williams et al.
competinginterests: |
  The authors declare no competing interests.
# OPTIONAL:
algorithms: true
# See https://publications.copernicus.org/for_authors/licence_and_copyright.html, normally used for transferring the copyright, if needed. 
# Note: additional copyright statements for affiliated software or data need to be placed in the data availability section. 
copyrightstatement: |
### The author's copyright for this publication is transferred to institution/company. 
### Note: unless stated otherwise, software and data affiliated with the manuscript are assumed to be published under the same licence as the article (currently Creative Commons 4.0)
availability:
  #  use this to add a statement when having only software code available
  #  use this to add a statement when having only data sets available
  #code: |
  #data: |
  codedata: |
    use this to add a statement when having data sets and software code available
authorcontribution: |
  (fill this in)
disclaimer: |
acknowledgements: |
  Thank you to all researchers who collected and published the data contained in ForC, and to all research assistants and collaborators who have helped to build the database.
  Funding for this study was provided by Bezos Earth Fund to The Nature Conservancy, the Institute for Global Environmental Strategies, WLS(?)
appendix: |
   \section{Mapping ForC to EFDB}

@@

Reflecting their strong influence on Earth's climate, forests play a central role in international plans for climate change mitigation under the Paris Agreement [@unfccc_adoption_2015].
Forest conservation, reforestation, and improved sustainable management all have significant -- and relatively cost-effective -- potential as climate change mitigation options, with conservation and reforestation having the fourth and fifth largest net emission reduction potentials or all mitigation options [@ipcc_summary_2022].
As of 2016, forest-based mitigation accounted for 26% of total planned greenhouse gas mitigation within Nationally Determined Contributions under the Paris Agreement [@grassi_key_2017].
Yet, envisioned forest-based climate change mitigation initiatives do not always correspond to actual emission reductions through on-the-ground implementation [e.g., @badgley_systematic_2022]<!--could be good to add some more references here-->.
One critical need for ensuring that forest-based climate change mitigation initiatives are effective is realistic planning and reporting, underlain by solid scientific data [@anderson-teixeira_effective_2022; @deng_comparing_2021].

The International Panel on Climate Change (IPCC) provides guidance for national greenhouse gas inventories for reporting to the United Nations Framework Convention on Climate Change [UNFCCC, @ipcc_2019_2019].
Under this guidance, greenhouse gas inventories include all managed land, including most of the world's forest land [@ogle_delineating_2018].
The IPCC inventory guidelines include specific instructions for accounting for greenhouse gas (mainly CO~2~) exchanges between forest land and the atmosphere [@ipcc_agriculture_2006; @ipcc_2019_2019].
This guidance has improved over the years as more of the relevant underlying data has become available [@requenasuarez_estimating_2019; @rozendaal_aboveground_2022], but there remains room for continuous improvement as the science advances.
For example, the year following the release of the latest IPCC guidelines, @cook-patton_mapping_2020 found that the latest default rates may underestimate rates of C accumulation in regrowth forests by 32% on average and fail to capture eight-fold variation within ecozones. 
In addition, @cuni-sanchez_high_2021 found that aboveground C stocks in mature African tropical montane forests were two-thirds higher than the IPCC default values for these forests.
This rapid evolution of scientific information on the climate mitigation potential of forests is beneficial to climate mitigation efforts, but requires improved mechanisms for communicating the latest information from scientific researchers to the practitioners who need reliable estimates for greenhouse gas mitigation planning. 
Moreover, high variability of forest C cycling within ecozones [e.g., @cook-patton_mapping_2020; @refs] implies that it is useful for practitioners to have access to locally-specific information, when available.

To improve the data accessible for C accounting, the IPCC created the Emission Factor Database (EFDB; https://www.ipcc-nggip.iges.or.jp/EFDB/main.php), which is intended as a recognized library of emission factors and other parameters that can be used for estimating greenhouse gas emissions and removals.
The EFDB is intended as a recognized library, and can be used both for efforts to tally a nation's intended or accomplished greenhouse gas reductions, or as a basis of comparison for external parties to evaluate these inventories. 
The EFDB encourages researchers to submit estimates of emission factors or other related parameters from peer-reviewed journal papers or other accepted sources for inclusion in the database. 
In the case of forests, emission factors refer to ... <!--C stocks and increments, and check other GHGs-->

### 2022-04-11 doc/manuscript/ESSDmanuscript.tex
work on intro

\copyrightstatement{The author's copyright for this publication is
transferred to institution/company.}

@@

Reflecting their strong influence on Earth's climate, forests play a
central role in international plans for climate change mitigation under
the Paris Agreement \citep{unfccc_adoption_2015}. Forest conservation,
reforestation, and improved sustainable management all have significant
-- and relatively cost-effective -- potential as climate change
mitigation options, with conservation and reforestation having the
fourth and fifth largest net emission reduction potentials or all
mitigation options \citep{ipcc_summary_2022}. As of 2016, forest-based
mitigation accounted for 26\% of total planned greenhouse gas mitigation
within Nationally Determined Contributions under the Paris Agreement
\citep{grassi_key_2017}. Yet, envisioned forest-based climate change
mitigation initiatives do not always correspond to actual emission
reductions through on-the-ground implementation \citep{refs}. One
critical need for ensuring that forest-based climate change mitigation
initiatives are effective is realistic planning, underlain by solid
scientific data \citep{anderson-teixeira_effective_2022}.

The International Panel on Climate Change (IPCC) provides guidance for
national greenhouse gas inventories for reporting to the United Nations
Framework Convention on Climate Change
\citep[UNFCCC,][]{ipcc_2019_2019}. Under this guidance, greenhouse gas
inventories include all managed land, including most of the world's
forest land \citep{ogle_delineating_2018}. The IPCC inventory guidelines
include specific instructions for accounting for greenhouse gas (mainly
CO\textsubscript{2}) exchanges between forest land and the atmosphere
\citep{ipcc_agriculture_2006, ipcc_2019_2019}. This guidance has
improved over the years as more of the relevant underlying data has
become available
\citep{requenasuarez_estimating_2019, rozendaal_aboveground_2022}, but
there remains room for continuous improvement as the science advances.
For example, the year following the release of the latest IPCC
guidelines, \citet{cook-patton_mapping_2020} found that the latest
default rates may underestimate rates of C accumulation in regrowth
forests by 32\% on average and fail to capture eight-fold variation
within ecozones. In addition, \citet{cuni-sanchez_high_2021} found that
aboveground C stocks in mature African tropical montane forests were
two-thirds higher than the IPCC default values for these forests. This
rapid evolution of scientific information on the climate mitigation
potential of forests is beneficial to climate mitigation efforts, but
requires improved mechanisms for communicating the latest information
from scientific researchers to the practitioners who need reliable
estimates for greenhouse gas mitigation planning. Moreover, high
variability of forest C cycling within ecozones
\citep[e.g.,][]{cook-patton_mapping_2020, refs} implies that it is
useful for those compiling national greenhouse gas inventories to have
access to locally-specific information, when available. To improve the
data accessible for C accounting, the IPCC created the Emission Factor
Database (EFDB; https://www.ipcc-nggip.iges.or.jp/EFDB/main.php), which
is intended as a recognized library of emission factors and other
parameters that can be used for estimating greenhouse gas emissions and
removals.

\textbf{To ensure that planned emissions reductions are realistic,
high-quality estimates of forest C stocks and fluxes must be publicly
accessible.} The EFDB is intended as a recognized library, and can be
used both for efforts to tally a nation's intended or accomplished
greenhouse gas reductions, or as a basis of comparison for external
parties to evaluate these inventories.

>>>>

@@

Reflecting their strong influence on Earth's climate, forests play a
central role in international plans for climate change mitigation under
the Paris Agreement \citep{unfccc_adoption_2015}. Forest conservation,
reforestation, and improved sustainable management all have significant
-- and relatively cost-effective -- potential as climate change
mitigation options, with conservation and reforestation having the
fourth and fifth largest net emission reduction potentials or all
mitigation options \citep{ipcc_summary_2022}. As of 2016, forest-based
mitigation accounted for 26\% of total planned greenhouse gas mitigation
within Nationally Determined Contributions under the Paris Agreement
\citep{grassi_key_2017}. Yet, envisioned forest-based climate change
mitigation initiatives do not always correspond to actual emission
reductions through on-the-ground implementation
\citep[e.g.,][]{badgley_systematic_2022}. One critical need for ensuring
that forest-based climate change mitigation initiatives are effective is
realistic planning and reporting, underlain by solid scientific data
\citep{anderson-teixeira_effective_2022, deng_comparing_2021}.

The International Panel on Climate Change (IPCC) provides guidance for
national greenhouse gas inventories for reporting to the United Nations
Framework Convention on Climate Change
\citep[UNFCCC,][]{ipcc_2019_2019}. Under this guidance, greenhouse gas
inventories include all managed land, including most of the world's
forest land \citep{ogle_delineating_2018}. The IPCC inventory guidelines
include specific instructions for accounting for greenhouse gas (mainly
CO\textsubscript{2}) exchanges between forest land and the atmosphere
\citep{ipcc_agriculture_2006, ipcc_2019_2019}. This guidance has
improved over the years as more of the relevant underlying data has
become available
\citep{requenasuarez_estimating_2019, rozendaal_aboveground_2022}, but
there remains room for continuous improvement as the science advances.
For example, the year following the release of the latest IPCC
guidelines, \citet{cook-patton_mapping_2020} found that the latest
default rates may underestimate rates of C accumulation in regrowth
forests by 32\% on average and fail to capture eight-fold variation
within ecozones. In addition, \citet{cuni-sanchez_high_2021} found that
aboveground C stocks in mature African tropical montane forests were
two-thirds higher than the IPCC default values for these forests. This
rapid evolution of scientific information on the climate mitigation
potential of forests is beneficial to climate mitigation efforts, but
requires improved mechanisms for communicating the latest information
from scientific researchers to the practitioners who need reliable
estimates for greenhouse gas mitigation planning. Moreover, high
variability of forest C cycling within ecozones
\citep[e.g.,][]{cook-patton_mapping_2020, refs} implies that it is
useful for practitioners to have access to locally-specific information,
when available.

To improve the data accessible for C accounting, the IPCC created the
Emission Factor Database (EFDB;
https://www.ipcc-nggip.iges.or.jp/EFDB/main.php), which is intended as a
recognized library of emission factors and other parameters that can be
used for estimating greenhouse gas emissions and removals. The EFDB is
intended as a recognized library, and can be used both for efforts to
tally a nation's intended or accomplished greenhouse gas reductions, or
as a basis of comparison for external parties to evaluate these
inventories. The EFDB encourages researchers to submit estimates of
emission factors or other related parameters from peer-reviewed journal
papers or other accepted sources for inclusion in the database. In the
case of forests, emission factors refer to \ldots{}
